\begin{abstract}
Trained sparse coders --- Sparse Autoencoders (SAEs), adVersarial Parameter Decomposition (\vpd), and Block-Sparse Featurizers (\bsf) --- have become the default tools for interpretable decomposition of transformer weights and activations. We show that for the specific case of weight-matrix decomposition, a training-free method built on the singular value decomposition matches or dominates every trained baseline we tested, and reproduces the qualitative findings that motivate them without any training. Given a weight matrix $W = U\Sigma V^\top$, our method (\svdomp) uses the SVD as a deterministic orthogonal atomic dictionary and selects atoms per input by the closed-form score $s_c(\phi) = \sigma_c |v_c^\top \phi|$. On Goodfire's public 67M \textsc{LlamaSimpleMLP}, \svdomp\ Pareto-dominates \vpd\ on 18 of 24 weight matrices, wins 24 / 24 on sparse reconstruction, faithfulness, and coherence, and 18 / 24 on support stability; the six losses are exactly the modules where the Davis-Kahan theorem predicts degradation. The dominance is not confined to reconstruction: on causal-ablation faithfulness --- how irreplaceable each selected component is when ablated --- \svdomp's per-input supports beat \vpd's trained gate 24 / 24, by $7$--$8\times$, because \vpd's gate collapses to a single input-independent support. \bsf's headline finding that vision concepts are 2--4-dimensional reproduces without training on real language-model activations: analytic \svdomp\ blocks track \bsf-warm to two decimal places at every block size. Trained decomposition can beat analytic only when the objective escapes Eckart-Young: on adversarial non-Frobenius constructions we recover 73.9\% of downstream MSE with a small trainable correction on top of the SVD scaffold. We interpret these results as evidence that trained interpretability probes on weight matrices are approximating a quantity that classical linear algebra already provides. A 42-test synthetic suite reproduces every headline claim on CPU in under a minute.
\end{abstract}
