\section{Related Work}
\label{sec:related}

\paragraph{Sparse dictionary learning for interpretability.} Sparse Autoencoders decompose neural activations into learned dictionary atoms with $L_1$ or Top-K sparsity \citep{bricken2023towards, cunningham2023sparse, gao2024topksae}. \vpd\ \citep{bushnaq2026vpd} adapts the SAE framework to weight decomposition by learning $V, U$ and a stochastic importance gate. \bsf\ \citep{goodfire2026bsf} generalizes to block-level sparsity with Grassmannian, block, and group-lasso variants, and reports 2--4-dimensional concept structure on vision. Our contribution is not a new sparse-coding method but the observation that the SVD supplies a Frobenius-optimal decomposition of a weight matrix analytically, and empirically dominates the trained methods on their own metrics; the block extension inherits the same property.

\paragraph{Jacobian and linearization-based interpretability.} Recent work on Jacobian lenses \citep{anthropic2026jlens} computes a linearized projection of activations onto vocabulary directions and shows that a substantial fraction of a language model's "reportable content" lives in a low-dimensional subspace of that linearization. This shares our observation that analytic linearizations often approximate what trained probes are learning; \svdomp\ is the weight-side analog of the same move. The Logit Lens \citep{nostalgebraist2020logitlens} and Tuned Lens \citep{belrose2023tunedlens} are related activation-space probes that use per-layer linear readouts.

\paragraph{Classical SVD, matching pursuit, and compressed sensing.} The Eckart-Young theorem \citep{eckart1936approximation} establishes Frobenius optimality of top-$k$ SVD; Weyl and Davis-Kahan bounds \citep{davis1970rotation} give quantitative stability. Orthogonal matching pursuit \citep{pati1993omp} is a greedy sparse recovery algorithm with approximate guarantees under RIP conditions \citep{tropp2004greed}; when the dictionary is orthogonal, OMP collapses to closed-form top-$k$. Prior work in compressed sensing has used SVD dictionaries for image compression and linear inverse problems \citep{aharon2006ksvd}; our contribution is the application to interpretability of transformer weight matrices, and the empirical demonstration that this analytic approach dominates learned dictionaries on this task.

\paragraph{Model compression via SVD.} Low-rank factorization of transformer weights for compression \citep{hsu2022svd, kaushal2023lord} is a related but distinct thread: those methods aim to reduce parameter count while preserving downstream task quality, not to produce an interpretable per-input decomposition. \svdomp\ uses the same primitive (SVD of $W$) for a different downstream objective (per-input sparse selection).
