\section{Discussion}
\label{sec:discussion}

\paragraph{Substrate matters.} All our results concern \emph{weight} decomposition. \bsf's original result is on \emph{activation} decomposition. Language-model activations are known to be non-aligned with principal directions of variance in practice (this is why SAEs beat PCA baselines on interpretability metrics), so we do not claim \svdomp\ would beat \bsf\ on activations. What we do claim: on weight decomposition, the analytic SVD dictionary dominates trained baselines; and the stable-rank plateau that \bsf\ interprets as "concepts are 2--4-dimensional" is method-independent, because analytic blocks reproduce it exactly.

\paragraph{Efficiency angle.} A concurrent benefit of \svdomp\ is that it eliminates the training compute of per-matrix probes. Table~\ref{tab:wall-clock} reports mean per-matrix wall-clock on both CPU and a Modal-hosted T4 GPU. \svdomp\ prepares each matrix in \textbf{$\sim$102 ms on CPU} and $\sim$111 ms on GPU; the trained methods take 15$\times$ longer on CPU and 4$\times$ longer on GPU. For interpretability at reasoning-model scale (hundreds of layers, thousands of interpreted modules), one SVD per module becomes a linear-algebra call at load time while a trained probe per module becomes a substantial infrastructure investment. We view this as a favorable side effect rather than the primary contribution.

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrrr}
\toprule
& \multicolumn{2}{c}{CPU} & \multicolumn{2}{c}{T4 GPU} \\
Method & ms & ratio & ms & ratio \\
\midrule
\svdomp\ (analytic)              & 102  & 1.0$\times$  & 111  & 1.0$\times$ \\
Block-\svdomp\ (analytic)        & 94   & 0.9$\times$  & 96   & 0.9$\times$ \\
Scaffold-trainable ($2K$ params) & 392  & 3.8$\times$  & 266  & 2.4$\times$ \\
\vpd\ (trained)                  & 1585 & 15.5$\times$ & 443  & 4.0$\times$ \\
\bsf-W cold                      & 1493 & 14.6$\times$ & 429  & 3.9$\times$ \\
\bsf-W warm                      & 1582 & 15.5$\times$ & 526  & 4.7$\times$ \\
\bottomrule
\end{tabular}
\caption{Wall-clock preparation cost per matrix on both CPU and Modal-hosted T4 GPU, averaged over the 24 target modules of the 67M \textsc{LlamaSimpleMLP}. Trained baselines use 40 AdamW steps; scaffold-trainable freezes $V, U$ at the SVD. \svdomp\ is $\sim$15$\times$ faster on CPU and $\sim$4$\times$ faster on GPU than any trained method. Trained baselines benefit more from GPU acceleration but do not close the gap.}
\label{tab:wall-clock}
\end{table}

\paragraph{Where trained probes should be used.} Two regimes remain where trained probes are appropriate on weight matrices: (i) non-Frobenius objectives that compose $W$ with downstream nonlinearities and matrices, where our scaffold-mode trainable correction (2 parameters per block) captures 73.9\% of the achievable gain relative to analytic; and (ii) the six specific modules with compressed singular spectra where Davis-Kahan predicts \svdomp\ instability. For these cases, the SVD is the correct \emph{starting point}, not a competitor.

\paragraph{Limitations.} (i) All Goodfire-67M experiments use a small pretrained checkpoint. The Pareto dominance and the stable-rank plateau reproduce on real Pythia-70M activations, but a systematic study on modern reasoning-scale models remains future work; the efficiency argument suggests it is precisely at large scale that \svdomp's zero training cost becomes materially valuable. (ii) The non-Frobenius escape (Sec.\ \ref{sec:non-frobenius-exp}) uses adversarial synthetic $W_\text{next}$; on real neighboring-layer matrices the gain is smaller because $W_\text{next}$ has less structured misalignment with $W$'s SVD, though it remains positive on 24 / 24 modules.

\paragraph{Implications for alignment.} If trained interpretability probes on weight matrices are approximating a quantity that the SVD already provides, three implications follow for the alignment research program. First, when interpretability results depend on probe-specific claims (e.g., "the trained probe found this concept"), those claims should be checked against the analytic SVD baseline before being credited to the training procedure. Second, the actual novelty of a trained probe is what it adds on top of the analytic baseline: the correction, not the base decomposition. Third, at reasoning-model scales, interpretability infrastructure that requires no training per layer becomes tractable in ways that trained probes do not.

\paragraph{Conclusion.} On weight decomposition of a widely used interpretability benchmark model, the SVD dictionary provides a Frobenius-optimal per-input sparse decomposition that dominates trained baselines on 24 of 24 per-metric comparisons and 18 of 24 modules Pareto-overall. \bsf's cross-modal 2--4-dimensional concept finding reproduces on real language-model activations without training. Trained decomposition adds value on non-Frobenius composed objectives; our small trainable correction on top of the SVD scaffold captures 73.9\% of that value. Together, these results position analytic decomposition as the appropriate baseline for interpretability of large neural network weights, and provide a principled account of the regime where training legitimately adds signal.
