\section{Introduction}

Modern reasoning systems achieve their capability through deep transformer stacks whose weight matrices resist inspection at scale. Sparse Autoencoders \citep{bricken2023towards,cunningham2023sparse} and its successors (\vpd, \citealp{bushnaq2026vpd}; \bsf, \citealp{goodfire2026bsf}) provide interpretable decompositions of one layer's activations or weights by training an encoder-decoder pair per target. As reasoning models scale toward hundreds of layers, training a probe per layer becomes a nontrivial fraction of the interpretability budget. For deployment of interpretable reasoning systems, cost per layer directly gates how much of the model can be understood.

We ask whether this training cost is necessary for the weight-decomposition case. Our answer is that it is not. Given a weight matrix $W \in \R^{d_\text{out} \times d_\text{in}}$ with singular value decomposition $W = U\Sigma V^\top$, the SVD supplies a deterministic orthogonal dictionary of rank-1 atoms $\{\sigma_c u_c v_c^\top\}$ that is Frobenius-optimal at every truncation level by the Eckart-Young theorem \citep{eckart1936approximation}. Because these atoms are orthogonal, orthogonal matching pursuit \citep{pati1993omp} on this dictionary collapses to a closed-form top-$k$ selection: per input $\phi$, score atom $c$ by
\begin{equation}
s_c(\phi) = \sigma_c \, |v_c^\top \phi|,
\label{eq:score}
\end{equation}
and select the top-$k$ atoms. No training, no random initialization, no learned parameters. We refer to this method as \svdomp.

\paragraph{Contributions.} This paper makes four contributions:

\begin{enumerate}[leftmargin=1.5em,itemsep=0.1em,topsep=0.2em]
\item \textbf{Training-free interpretable decomposition (Sec. \ref{sec:method}).} We define \svdomp\ (Sec. \ref{sec:svdomp}) and its block extension (Sec. \ref{sec:block}), and give closed-form selection rules for both.

\item \textbf{Efficiency analysis (Sec. \ref{sec:efficiency}).} We compare \svdomp\ against \vpd\ and \bsf\ on training compute, inference latency, and storage. \svdomp\ pays no training cost and reduces per-inference selection to a matmul and top-$k$.

\item \textbf{Empirical dominance on weight decomposition (Sec. \ref{sec:experiments}).} On Goodfire's 67M \textsc{LlamaSimpleMLP}, \svdomp\ Pareto-dominates \vpd\ on 18 of 24 target modules and wins every per-metric comparison 24/24 on sparse reconstruction, faithfulness, and coherence. A block version of \bsf\ warm-started from the SVD strictly beats random-init \bsf-W under identical training budget on 24 of 24 modules --- evidence that initialization, not the trained architecture, is the causal factor.

\item \textbf{Cross-modality replication of \bsf's stable-rank plateau (Sec. \ref{sec:stable-rank}).} \bsf\ reports concept dimensionality plateauing at $\sim$4 regardless of block size $K$ on vision activations. We reproduce the plateau on real Pythia-70M language-model activations at $\sim$2.1, using analytic \svdomp\ blocks. Both trained \bsf-W (warm-started) and analytic \svdomp\ blocks converge to identical stable ranks (2.10 vs.\ 2.10 at $K=16$). This shows the finding is a property of the activation distribution, not of the featurizer.
\end{enumerate}

\paragraph{Scope.} \svdomp\ dominates trained decompositions on Frobenius objectives on weights because Eckart-Young caps every rank-$k$ approximator at the truncated SVD. It does not dominate on \emph{non-Frobenius} objectives such as reconstruction after a nonlinear downstream layer, where training can rotate the dictionary toward downstream-important directions. We characterize this regime (Sec. \ref{sec:non-frobenius}) and give a trainable extension that escapes the Eckart-Young cap. On adversarial constructions the trainable extension achieves 73.9\% mean MSE reduction with a small parameter overhead.
